---
title: "Team Project Group 18"
author: "Nathan Patton, Reed Baumgardner, Fadumo Hussein"
output: 
 html_document:
  toc: true
  toc_depth: 3
  theme: lumen
  code_folding: show
---

# Executive Summary
One of the most unfortunate and disliked aspect of being an athlete is the risk of getting injured. Thousands of athletes are injured every single year, and these injuries vary with severity. In this report, the plan is to research how different variables effect the severity of injury in football. The end goal of this analysis is to lower the risk of injury for athletes, and make their lives healthier and easier, along with other stakeholders involved, such as family, teammates, athletic trainers, strength and conditioning coaches, etc. The established data set is detailed below, with emphasis on cleaning and sub-setting variables that are pertinent to the questions, along with a data dictionary that lists all the contents. To get a better understanding of the data sets contents, exploratory data analysis was performed, in addition to the data descriptions. In order to answer the questions of interest, detailed charts were created. Based on these charts, the key findings were.... 

# Report Dependencies
```{r}
library(tidyverse)
library(ggplot2)
library(mice)
library(skimr)
library(pollster)
library(kableExtra)
library(cowplot)
library(datadictionary)
library(pastecs)
```

# Establishing Data Set
[Kaggle-- NFL Injury Analysis](https://www.kaggle.com/code/aleksandradeis/nfl-injury-analysis/data)
```{r}
#Loading in the data from local files
InjuryRecord <- read.csv("InjuryRecord.csv")
PlayList <- read.csv("PlayList.csv")

#Cleaning Data
##Checking for missing values injury record 
InjuryRecord[InjuryRecord==""] <- NA
colSums(is.na(InjuryRecord))
md.pattern(InjuryRecord, rotate.names = TRUE)

##Checking for missing values playlist 
PlayList[PlayList==""] <- NA
colSums(is.na(PlayList))
md.pattern(PlayList, rotate.names = TRUE)

##Joining PlayList with InjuryRecord 
injuries <- inner_join(PlayList, InjuryRecord, by="PlayKey")

##Removing duplicate columns
injuries <- select(injuries, -c("PlayerKey.y","GameID.y","Surface", "Position", "PositionGroup"))

##Renaming Columns
injuries <- injuries %>%
  rename("PlayerKey" = "PlayerKey.x", "GameID" = "GameID.x")

##Temperatures for indoor stadiums were incorrectly showing up as -999. We set it to 70 degrees based on the average on-field temperature NFL teams keep indoor stadiums. 
injuries$Temperature[injuries$Temperature == -999] <- 70

##Weather for indoor stadiums was set as "No Weather" for the purposes of this project
injuries$Weather[is.na(injuries$Weather) == TRUE] <- "No Weather"

##Random PlayerDay was changed from -34 to 34
injuries$PlayerDay[injuries$PlayerDay==-34] <- 34

##Combining some of the weather values
injuries$Weather[injuries$Weather==""|
                   injuries$Weather=="Indoors"|
                   injuries$Weather=="Controlled Climate"]<- "Indoor"
injuries$Weather[injuries$Weather=="Clear skies"|
                   injuries$Weather=="Clear and warm"|
                   injuries$Weather=="Clear Skies"]<- "Clear"
injuries$Weather[injuries$Weather=="Coudy"|
                   injuries$Weather=="Cloudy and Cool"|
                   injuries$Weather=="Cloudy, 50% change of rain"|
                   injuries$Weather=="Cloudy with periods of rain, thunder possible. Winds shifting to WNW, 10-20 mph."|
                   injuries$Weather=="Mostly cloudy"|
                   injuries$Weather=="Partly Cloudy"] <- "Cloudy"
injuries$Weather[injuries$Weather=="Light Rain"|
                   injuries$Weather=="Rain shower"]<- "Rain"
injuries$Weather[injuries$Weather=="Mostly Sunny"|
                   injuries$Weather=="Mostly sunny"|
                   injuries$Weather=="Sun & clouds"]<- "Sunny"

##Combining PlayType Variables
injuries$PlayType[injuries$PlayType=="Kickoff Not Returned"|
                    injuries$PlayType=="Kickoff Returned"] <- "Kickoff"
injuries$PlayType[injuries$PlayType=="Punt Not Returned"|
                    injuries$PlayType=="Punt Returned"]<- "Punt"
injuries <- na.omit(injuries)

##Forming a new days missed column using mutate 
injuries <- injuries %>%
  mutate(Days_Missed = ifelse(DM_M42 == 1, 42, ifelse(DM_M28 == 1, 28, ifelse(DM_M7 == 1, 7, 1))))

##Deleting the original days missed columns
injuries <- select(injuries, -c("DM_M1","DM_M7","DM_M28","DM_M42"))
View(injuries)

#Subsetting variables pertinent to questions and creating data dictionaries
question1data <- injuries %>%
  select(RosterPosition, BodyPart, StadiumType, Temperature, Weather, PlayType, Days_Missed)
create_dictionary(question1data)
question2data <- question1data %>%
  filter(RosterPosition == "Wide Receiver")
create_dictionary(question2data)
```

#Background and Questions
## Background

## Questions
1. Which conditions are linked to the most and most severe injuries? 
2. Which conditions are most/least linked to WR injuries?

# Descriptive Statistics and Exploratory Data Analysis
## Descriptive Statistics
```{r}
summary(injuries)
```

## Exploratory Data Analysis
###Subsetting Variables Pertinent to Questions
```{r}
#1. Which conditions are linked to the most and most severe injuries? 
question1data <- injuries %>%
  select(RosterPosition, BodyPart, StadiumType, Temperature, Weather, PlayType, Days_Missed)
View(question1data)
summary(question1data)

#2. Which conditions are most/least linked to WR injuries?
question2data <- question1data %>%
  filter(RosterPosition == "Wide Receiver")
View(question2data)
summary(question2data)
```

### Checking Distributions of Continuous Variables
#### Predictor Variable: Temperature
```{r}
ggplot(injuries, aes(Temperature)) +
  geom_histogram(bins = 200, fill = '#540d6e') + 
  geom_vline(aes(xintercept = mean(Temperature)),col = 'lightgray', linewidth = 1) +
  geom_vline(aes(xintercept = median(Temperature)), col = 'lightgray', linetype = 'dashed') +
  labs(
    title = 'Histogram of Temperature',
    caption = 'mean: solid line, median = dashed line'
  ) +
  theme_minimal()
```
# Color Palette 
pal = c("#8CB369","#2191FB","#F4A259","#BC4B51")

# Question 1 Plots and Narrative
## Grouped Bar Chart for BodyPart by Days_Missed
```{r}
ggplot(question1data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
    xlab("Body Part Injured") +
    ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for FieldType by Days_Missed
```{r}
ggplot(question1data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Weather by Days_Missed
```{r}
ggplot(question1data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather Conditions") +
  xlab("Weather Conditions") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Position by Days_Missed
```{r}
ggplot(question1data, aes(x=RosterPosition, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Position") +
  xlab("Roster Position") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Play Type by Days_Missed
```{r}
ggplot(question1data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Combined Chart
```{r}
q1grid <- plot_grid(a1,a2,a3,a5,a4, 
          labels = c("A", "B", "C","D","E"),
          ncol = 2, nrow = 3)
q1grid
```


# Question 2 Plots and Narrative
## Grouped Bar Chart for BodyPart by Days_Missed

```{r}
b1<-ggplot(question2data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
  xlab("Body Part Injured") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for FieldType by Days_Missed
```{r}
b2<-ggplot(question2data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Weather by Days_Missed
```{r}
b3<-ggplot(question2data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather Conditions") +
  xlab("Weather Conditions") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Play Type by Days_Missed
```{r}
b4<-ggplot(question2data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Combined Chart
```{r}
q2grid <- plot_grid(b1,b2,b3,b4, 
                    labels = c("A", "B", "C","D"),
                    ncol = 2, nrow = 2)
q2grid
```
# Summary

# Citations
[STHDA (help with plot_grid())](http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/)
# Contrubutions:
Nathan Patton - Renaming columns and merging tables, cleaning, mutating days missed columns into one, and the first plot
Reed Baumgardner - Merging tables, removing duplicate columns and renaming columns, cleaning, additional comments on process
Fadumo Hussein - Find and relabel missing data, collaping variable factors, additional comments on process
