---
title: "Team Project Group 18"
author: "Nathan Patton, Reed Baumgardner, Fadumo Hussein"
output: 
 html_document:
  toc: true
  toc_depth: 3
  theme: lumen
  code_folding: show
---
# Executive Summary
This document outlines the 

# Report Dependencies
```{r}
library(tidyverse)
library(ggplot2)
library(mice)
library(skimr)
library(pollster)
library(kableExtra)
library(cowplot)
```

# Establishing Data Set
[Kaggle-- NFL Injury Analysis](https://www.kaggle.com/code/aleksandradeis/nfl-injury-analysis/data)
```{r}
#Loading in the data from local files
InjuryRecord <- read.csv("InjuryRecord.csv")
PlayList <- read.csv("PlayList.csv")

#Cleaning Data
##Checking for missing values injury record 
InjuryRecord[InjuryRecord==""] <- NA
colSums(is.na(InjuryRecord))
md.pattern(InjuryRecord, rotate.names = TRUE)

##Checking for missing values playlist 
PlayList[PlayList==""] <- NA
colSums(is.na(PlayList))
md.pattern(PlayList, rotate.names = TRUE)

##Joining PlayList with InjuryRecord 
injuries <- inner_join(PlayList, InjuryRecord, by="PlayKey")

##Removing duplicate columns
injuries <- select(injuries, -c("PlayerKey.y","GameID.y","Surface", "Position", "PositionGroup"))

##Renaming Columns
injuries <- injuries %>%
  rename("PlayerKey" = "PlayerKey.x", "GameID" = "GameID.x")

##Temperatures for indoor stadiums were incorrectly showing up as -999. We set it to 70 degrees based on the average on-field temperature NFL teams keep indoor stadiums. 
injuries$Temperature[injuries$Temperature == -999] <- 70

##Weather for indoor stadiums was set as "No Weather" for the purposes of this project
injuries$Weather[is.na(injuries$Weather) == TRUE] <- "No Weather"

##Random PlayerDay was changed from -34 to 34
injuries$PlayerDay[injuries$PlayerDay==-34] <- 34

##Combining some of the weather values
injuries$Weather[injuries$Weather==""|
                   injuries$Weather=="Indoors"|
                   injuries$Weather=="Controlled Climate"]<- "Indoor"
injuries$Weather[injuries$Weather=="Clear skies"|
                   injuries$Weather=="Clear and warm"|
                   injuries$Weather=="Clear Skies"]<- "Clear"
injuries$Weather[injuries$Weather=="Coudy"|
                   injuries$Weather=="Cloudy and Cool"|
                   injuries$Weather=="Cloudy, 50% change of rain"|
                   injuries$Weather=="Cloudy with periods of rain, thunder possible. Winds shifting to WNW, 10-20 mph."|
                   injuries$Weather=="Mostly cloudy"|
                   injuries$Weather=="Partly Cloudy"] <- "Cloudy"
injuries$Weather[injuries$Weather=="Light Rain"|
                   injuries$Weather=="Rain shower"]<- "Rain"
injuries$Weather[injuries$Weather=="Mostly Sunny"|
                   injuries$Weather=="Mostly sunny"|
                   injuries$Weather=="Sun & clouds"]<- "Sunny"

##Combining PlayType Variables
injuries$PlayType[injuries$PlayType=="Kickoff Not Returned"|
                    injuries$PlayType=="Kickoff Returned"] <- "Kickoff"
injuries$PlayType[injuries$PlayType=="Punt Not Returned"|
                    injuries$PlayType=="Punt Returned"]<- "Punt"

injuries <- na.omit(injuries)

##Forming a new days missed column using mutate 
injuries <- injuries %>%
  mutate(Days_Missed = ifelse(DM_M42 == 1, 42, ifelse(DM_M28 == 1, 28, ifelse(DM_M7 == 1, 7, 1))))

##Deleting the original days missed columns
injuries <- select(injuries, -c("DM_M1","DM_M7","DM_M28","DM_M42"))
View(injuries)
```

#Background and Questions
## Background

## Questions
1. Which conditions are linked to the most and most severe injuries? 
2. Which conditions are most/least linked to WR injuries?

# Descriptive Statistics and Exploratory Data Analysis
## Descriptive Statistics
```{r}
summary(injuries)
```

## Exploratory Data Analysis
###Subsetting Variables Pertinent to Questions
```{r}
#1. Which conditions are linked to the most and most severe injuries? 
question1data <- injuries %>%
  select(RosterPosition, BodyPart, StadiumType, Temperature, Weather, PlayType, Days_Missed)
View(question1data)

#2. Which conditions are most/least linked to WR injuries?
question2data <- question1data %>%
  filter(RosterPosition == "Wide Receiver")
View(question2data)
```

### Checking Distributions of Continuous Variables
#### Predictor Variable: Temperature
```{r}
ggplot(injuries, aes(Temperature)) +
  geom_histogram(bins = 200, fill = '#540d6e') + 
  geom_vline(aes(xintercept = mean(Temperature)),col = 'lightgray', linewidth = 1) +
  geom_vline(aes(xintercept = median(Temperature)), col = 'lightgray', linetype = 'dashed') +
  labs(
    title = 'Histogram of Temperature',
    caption = 'mean: solid line, median = dashed line'
  ) +
  theme_minimal()
```

# Question 1 Plots and Narrative
## Grouped Bar Chart for BodyPart by Days_Missed
```{r}
ggplot(question1data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
    xlab("Body Part Injured") +
    ylab("Count")
```

## Grouped Bar Chart for FieldType by Days_Missed
```{r}
ggplot(question1data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")
```

## Grouped Bar Chart for Weather by Days_Missed
```{r}
ggplot(question1data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather Conditions") +
  xlab("Weather Conditions") +
  ylab("Count")
```

## Grouped Bar Chart for Position by Days_Missed
```{r}
ggplot(question1data, aes(x=RosterPosition, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Position") +
  xlab("Roster Position") +
  ylab("Count")
```

## Grouped Bar Chart for Play Type by Days_Missed
```{r}
ggplot(question1data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")
```

## Combined Chart
```{r}
q1grid <- plot_grid(a1,a2,a3,a5,a4, 
          labels = c("A", "B", "C","D","E"),
          ncol = 2, nrow = 3)
q1grid
```


# Question 2 Plots and Narrative
## Grouped Bar Chart for BodyPart by Days_Missed

```{r}
b1<-ggplot(question2data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
  xlab("Body Part Injured") +
  ylab("Count")
```

## Grouped Bar Chart for FieldType by Days_Missed
```{r}
b2<-ggplot(question2data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")
```

## Grouped Bar Chart for Weather by Days_Missed
```{r}
b3<-ggplot(question2data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather Conditions") +
  xlab("Weather Conditions") +
  ylab("Count")
```

## Grouped Bar Chart for Play Type by Days_Missed
```{r}
b4<-ggplot(question2data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")
```

## Combined Chart
```{r}
q2grid <- plot_grid(b1,b2,b3,b4, 
                    labels = c("A", "B", "C","D"),
                    ncol = 2, nrow = 2)
q2grid
```

# Contrubutions:
Nathan Patton - Renaming columns and merging tables, cleaning, mutating days missed columns into one, and the first plot
Reed Baumgardner - Merging tables, removing duplicate columns and renaming columns, cleaning, additional comments on process
Fadumo Hussein - Find and relabel missing data, collaping variable factors, additional comments on process
