---
title: "Team Project Group 18"
output: 
 html_document:
  toc: true
  toc_depth: 3
  theme: lumen
  code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Group Names
Reed Baumgardner
Nathan Patton
Fadumo Hussein

# Topic
NFL Injuries

# Possible Data Sources
[Kaggle-- NFL Injury Analysis](https://www.kaggle.com/code/aleksandradeis/nfl-injury-analysis/data)

# Possible Questions
What variables correlate positively with NFL injuries? (injury surface, weather, stadium type, position, etc)

Is it more common to get injured in the cold/rain? Indoor/Outdoor stadiums?

Which variables are causing the most severe injuries (measured by days out)?

Which position is most at risk? In which conditions?

Analysis per position. Which variables are linked to the most QB injuries?

#Final Questions
1. Which conditions are linked to the most and most severe injuries? 
2. Which conditions are most/least linked to QB injuries?

#Packages 
```{r}
library(tidyverse)
library(ggplot2)
library(mice)
library(skimr)
library(pollster)
library(kableExtra)
```

#Uplaod Data 
```{r}
InjuryRecord <- read.csv("InjuryRecord.csv")
PlayList <- read.csv("PlayList.csv")
```

#Checking for missing values and cleaning up data 
```{r}
#Checking for missing values injury record 
InjuryRecord[InjuryRecord==""] <- NA
colSums(is.na(InjuryRecord))
md.pattern(InjuryRecord, rotate.names = TRUE)

#Checking for missing values playlist 
PlayList[PlayList==""] <- NA
colSums(is.na(PlayList))
md.pattern(PlayList, rotate.names = TRUE)
```

#Rename columns and merging table 
```{r}
#Joining PlayList with InjuryRecord 
injuries <- inner_join(PlayList, InjuryRecord, by="PlayKey")

#Removing duplicate columns
injuries <- select(injuries, -c("PlayerKey.y","GameID.y","Surface", "Position", "PositionGroup"))

#Renaming Columns
injuries <- injuries %>%
  rename("PlayerKey" = "PlayerKey.x", "GameID" = "GameID.x")
```

#More Cleaning
```{r}
# Temperatures for indoor stadiums were incrrectly showinf up as -999. We set it to 70 degrees based on the average on-field temperature NFL teams keep indoor stadiums. 
injuries$Temperature[injuries$Temperature == -999] <- 70
# Weather for indoor stadiums was set as "No Weather" for the purposes of this project
# Cleaning Weather responses into repeatable responses and fixing other errors/NAs.
injuries$Weather[is.na(injuries$Weather) == TRUE] <- "No Weather"
injuries$PlayerDay[injuries$PlayerDay==-34]<-34
injuries$Weather[injuries$Weather==""]<- "Indoor"
injuries$Weather[injuries$Weather=="Indoors"]<- "Indoor"
injuries$Weather[injuries$Weather=="Controlled Climate"]<- "Indoor"
injuries$Weather[injuries$Weather=="Clear skies"|
                   injuries$Weather=="Clear and warm"|
                   injuries$Weather=="Clear Skies"|
                   injuries$Weather=="Fair"|
                   injuries$Weather == "Cold"]<- "Clear"
injuries$Weather[injuries$Weather=="Coudy"|
                   injuries$Weather=="Cloudy and Cool"|
                   injuries$Weather=="Cloudy, 50% change of rain"|
                   injuries$Weather=="Cloudy with periods of rain, thunder possible. Winds shifting to WNW, 10-20 mph."|
                   injuries$Weather=="Mostly cloudy" | 
                   injuries$Weather=="Partly Cloudy"] <- "Cloudy"
injuries$Weather[injuries$Weather=="Light Rain"|
                   injuries$Weather=="Rain shower"]<- "Rain"
injuries$Weather[injuries$Weather=="Mostly Sunny"|
                   injuries$Weather=="Mostly sunny"|
                   injuries$Weather=="Sun & clouds"]<- "Sunny"

View(injuries)
```

#Merging days missed columns into one outcome column
```{r}
#Forming a new days missed column using mutate 
injuries <- injuries %>%
  mutate(Days_Missed = ifelse(DM_M42 == 1, 42, ifelse(DM_M28 == 1, 28, ifelse(DM_M7 == 1, 7, 1))))

#Deleting the original days missed columns
injuries <- select(injuries, -c("DM_M1","DM_M7","DM_M28","DM_M42"))
View(injuries)
```

#First Plot
```{r}
ggplot(injuries,               
       aes(x = Temperature,           
           y = Days_Missed,        
           color = RosterPosition)) +   
  geom_point(aes(shape = FieldType))
```
#Contrubutions:
Nathan Patton - Renaming columns and merging tables, cleaning, mutating days missed columns into one, and the first plot
Reed Baumgardner - Merging tables, removing duplicate columns and renaming columns, cleaning, additional comments on process
Fadumo Hussein - Find and relabel missing data, collaping variable factors, additional comments on process




#Exploratory Data Analysis
##Subsetting Variables Pertinent to Questions
###1. Which conditions are linked to the most and most severe injuries?

question1data <- injuries %>%
  select(RosterPosition, FieldType, Temperature, Weather, PlayType, Days_Missed)


### Days_Missed as factor variable for grouped bar charts purposes (skip if not needed)
question1data$Days_Missed <- as.factor(question1data$Days_Missed)

View(question1data)

##Checking Distributions of Continuous Variables
###Outcome Variables - Number of Games Missed (Need to do some editing and pivot longer before doing this)
###Predictor Variables
####Temperature

ggplot(injuries, aes(Temperature)) +
  geom_histogram(bins = 200, fill = '#540d6e') + 
  geom_vline(aes(xintercept = mean(Temperature)),col = 'lightgray', linewidth = 1) +
  geom_vline(aes(xintercept = median(Temperature)), col = 'lightgray', linetype = 'dashed') +
  labs(
    title = 'Histogram of Temperature',
    caption = 'mean: solid line, median = dashed line'
  ) +
  theme_minimal()


##Checking for Outliers in Continuous Variables

ggplot(injuries, aes(Temperature)) +
  geom_boxplot(fill = '#540d6e') + 
  labs(
    title = 'Boxplot of Temperature'
  ) +
  theme_minimal()


# Grouped Bar Charts
## Grouped Bar Chart for BodyPart by Days_Missed
ggplot(question1data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
    xlab("Body Part Injured") +
    ylab("Count")

## Grouped Bar Chart for FieldType by Days_Missed

ggplot(question1data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")

## Grouped Bar Chart for Weather by Days_Missed

ggplot(question1data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather Conditions") +
  xlab("Weather Conditions") +
  ylab("Count")

## Grouped Bar Chart for Position by Days_Missed

ggplot(question1data, aes(x=RosterPosition, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Position") +
  xlab("Roster Position") +
  ylab("Count")

## Grouped Bar Chart for Play Type by Days_Missed

ggplot(question1data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")



# Link for multiple graphs on one page http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/

