---
title: "Team Project Group 18"
author: "Nathan Patton, Reed Baumgardner, Fadumo Hussein"
output: 
 html_document:
  toc: true
  toc_depth: 3
  theme: lumen
  code_folding: show
---
# Executive Summary
One of the most unfortunate and disliked aspect of being an athlete is the risk of getting injured. Thousands of athletes are injured every single year, and these injuries vary with severity. In this report, the plan is to research how different variables effect the severity of injury in football. The end goal of this analysis is to lower the risk of injury for athletes, and make their lives healthier and easier, along with other stakeholders involved, such as family, teammates, athletic trainers, strength and conditioning coaches, etc. The established data set is detailed below, with emphasis on cleaning and sub-setting variables that are pertinent to the questions, along with a data dictionary that lists all the contents. To get a better understanding of the data sets contents, exploratory data analysis was performed, in addition to the data descriptions. In order to answer the questions of interest, detailed charts were created. Based on these charts, the key findings were.... 

# Report Dependencies
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(mice)
library(skimr)
library(pollster)
library(kableExtra)
library(cowplot)
library(datadictionary)
library(pastecs)
```

# Establishing Data Set
[Kaggle-- NFL Injury Analysis](https://www.kaggle.com/code/aleksandradeis/nfl-injury-analysis/data)
```{r}
#Loading in the data from local files
InjuryRecord <- read.csv("InjuryRecord.csv")
PlayList <- read.csv("PlayList.csv")

#Cleaning Data
##Checking for missing values injury record 
InjuryRecord[InjuryRecord==""] <- NA
colSums(is.na(InjuryRecord))
md.pattern(InjuryRecord, rotate.names = TRUE)

##Checking for missing values playlist 
PlayList[PlayList==""] <- NA
colSums(is.na(PlayList))
md.pattern(PlayList, rotate.names = TRUE)

##Joining PlayList with InjuryRecord 
injuries <- inner_join(PlayList, InjuryRecord, by="PlayKey")

##Removing duplicate columns
injuries <- select(injuries, -c("PlayerKey.y","GameID.y","Surface", "Position", "PositionGroup"))

##Renaming Columns
injuries <- injuries %>%
  rename("PlayerKey" = "PlayerKey.x", "GameID" = "GameID.x")

##Temperatures for indoor stadiums were incorrectly showing up as -999. We set it to 70 degrees based on the average on-field temperature NFL teams keep indoor stadiums. 
injuries$Temperature[injuries$Temperature == -999] <- 70

##Weather for indoor stadiums was set as "Indoor" for the purposes of this project
injuries$Weather[is.na(injuries$Weather) == TRUE] <- "Indoor"

##Random PlayerDay was changed from -34 to 34
injuries$PlayerDay[injuries$PlayerDay==-34] <- 34

##Combining some of the weather values
injuries$Weather[injuries$Weather==""|
                   injuries$Weather=="Indoors"|
                   injuries$Weather=="Controlled Climate"]<- "Indoor"
injuries$Weather[injuries$Weather=="Clear skies"|
                   injuries$Weather=="Clear and warm"|
                   injuries$Weather=="Clear Skies"|
                   injuries$Weather=="Fair"|
                   injuries$Weather=="Cold"]<- "Clear"
injuries$Weather[injuries$Weather=="Coudy"|
                   injuries$Weather=="Cloudy and Cool"|
                   injuries$Weather=="Cloudy, 50% change of rain"|
                   injuries$Weather=="Cloudy with periods of rain, thunder possible. Winds shifting to WNW, 10-20 mph."|
                   injuries$Weather=="Mostly cloudy"|
                   injuries$Weather=="Partly Cloudy"] <- "Cloudy"
injuries$Weather[injuries$Weather=="Light Rain"|
                   injuries$Weather=="Rain shower"]<- "Rain"
injuries$Weather[injuries$Weather=="Mostly Sunny"|
                   injuries$Weather=="Mostly sunny"|
                   injuries$Weather=="Sun & clouds"]<- "Sunny"

##Combining PlayType Variables
injuries$PlayType[injuries$PlayType=="Kickoff Not Returned"|
                    injuries$PlayType=="Kickoff Returned"] <- "Kickoff"
injuries$PlayType[injuries$PlayType=="Punt Not Returned"|
                    injuries$PlayType=="Punt Returned"]<- "Punt"

injuries <- na.omit(injuries)

##Forming a new days missed column using mutate 
injuries <- injuries %>%
  mutate(Days_Missed = ifelse(DM_M42 == 1, 42, ifelse(DM_M28 == 1, 28, ifelse(DM_M7 == 1, 7, 1))))

##Deleting the original days missed columns
injuries <- select(injuries, -c("DM_M1","DM_M7","DM_M28","DM_M42"))

question1data <- injuries %>%
  select(RosterPosition, BodyPart, StadiumType, Temperature, Weather, PlayType, Days_Missed)

create_dictionary(question1data)

question2data <- question1data %>%
  filter(RosterPosition == "Wide Receiver")

create_dictionary(question2data)
```

## Background
This dataset was created by the National Football League (NFL) for a Kaggle-based analytics competition named "NFL 1st and Future - Analytics." The purpose of this competition was to "investigate the relationship between the playing surface and the injury and performance of NFL athletes." The inspiration for the competition was the recent investigations into the higher rate of injuries to NFL players on synthetic turf compared to that on natural turf. As a group, we were interested in sports and health, and this dataset was a reputable intersection of the two. In one of the nation's most profitable, yet dangerous, sports leagues, the risk of investment into a player can be lowered if steps continue to be taken to ensure player safety. While NFL player health is the foremost reason for the creation of this competition and dataset, we believe that conclusions from our analysis can be beneficial to the youth ranks of football as well. Keeping the children safe on the field was an important motivation for our analysis.  Our questions were stemmed from the NFL's own questions about safety. Our first question was an obvious one to come to. We wanted to explore all the variables in the dataset to find which ones were contributing most to injuries, especially severe ones that are career and sometimes life changing. Our second question was derived more specifically as we wanted to look into one of the NFL's most injury-prone positions-- Wide Receivers. This position group are some of the highest paid in the league and also one of the positions that are the most affected long-term by injuries as speed, agility, and sharpness are absolute keys to their success. These factors led us to focusing an analysis on just the wide receivers.

## Questions
1. Which conditions are linked to the most severe injuries? 
2. Which conditions are most/least linked to WR injuries?

# Descriptive Statistics and Exploratory Data Analysis
## Descriptive Statistics
```{r}
summary(injuries)
```

## Exploratory Data Analysis
###Subsetting Variables Pertinent to Questions
```{r}
#1. Which conditions are linked to the most and most severe injuries? 
question1data <- injuries %>%
  select(RosterPosition, BodyPart, StadiumType, Temperature, Weather, PlayType, Days_Missed, FieldType)

question1data$Days_Missed <- as.factor(question1data$Days_Missed)
View(question1data)

summary(question1data)

#2. Which conditions are most/least linked to WR injuries?
question2data <- question1data %>%
  filter(RosterPosition == "Wide Receiver")
question2data$Days_Missed <- as.factor(question2data$Days_Missed)

View(question2data)

summary(question2data)
```

### Checking Distributions of Continuous Variables
#### Predictor Variable: Temperature
```{r}
ggplot(injuries, aes(Temperature)) +
  geom_histogram(bins = 200, fill = '#540d6e') + 
  geom_vline(aes(xintercept = mean(Temperature)),col = 'lightgray', linewidth = 1) +
  geom_vline(aes(xintercept = median(Temperature)), col = 'lightgray', linetype = 'dashed') +
  labs(
    title = 'Histogram of Temperature',
    caption = 'mean: solid line, median = dashed line'
  ) +
  theme_minimal()
```

# Color Palette 
```{r}
pal = c("#8CB369","#2191FB","#F4A259","#BC4B51")
```

# Question 1 Plots and Narrative
## Grouped Bar Chart for BodyPart by Days_Missed
```{r}
a1<-ggplot(question1data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
    xlab("Body Part Injured") +
    ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for FieldType by Days_Missed
```{r}
a2<-ggplot(question1data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Weather by Days_Missed
```{r}
a3<-ggplot(question1data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather") +
  xlab("Weather Conditions") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45))
```


## Grouped Bar Chart for Play Type by Days_Missed
```{r}
a4<-ggplot(question1data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")+
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Combined Chart
```{r}
q1grid <- plot_grid(a1,a2,a3,a4, 
          labels = c("A", "B", "C","D"),
          ncol = 2, nrow = 2)
q1grid
```

# Question 2 Plots and Narrative
## Grouped Bar Chart for BodyPart by Days_Missed

```{r}
b1<-ggplot(question2data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
  xlab("Body Part Injured") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for FieldType by Days_Missed
```{r}
b2<-ggplot(question2data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Weather by Days_Missed
```{r}
b3<-ggplot(question2data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather") +
  xlab("Weather Conditions") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Play Type by Days_Missed
```{r}
b4<-ggplot(question2data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Combined Chart
```{r}
q2grid <- plot_grid(b1,b2,b3,b4, 
                    labels = c("A", "B", "C","D"),
                    ncol = 2, nrow = 2)
q2grid
```

## Narrative 

We focused our next question on wide receivers. Once again we wanted to focus on this group of athletes due to How injury prone this role is. Injuries to the ankle cause these athletes to sit out for the longest.  there was no difference in field types.  however similar to the first set of combined graphs,  play that took place in cloudy weather led to longer injury days. Finally, plays that were a pass also led to wide receivers being out the longest. 

As stayed similarly before knowing our stakeholders and their goal to protect the health and wealth of our players here are potential solutions for wide receivers. The first would be to reschedule football games if the weather is cloudy.  another solution can be creating harsher penalties.  Finally, we can be looking at Having wide receivers wear ankle braces to ensure the stability of their ankles. However, we do recognize that adding any potential weight to them will slow their speed down.

# Summary
Our goal in this report is to ultimately give insight as to how we can ensure the safety of athletes.  This was our intent in each step of this report whether it be data cleaning looking at data descriptives and building plots that can give us insight but also potential solutions on how to keep athletes safe. A major finding was cloudy days and passes both had very high days missed. Injuries were high as well however varied based on position. Overall looking at the data we feel confident that we were able to answer both questions and additionally provide potential solutions to limit the risk of injury for players in the NFL.

# Citations
[STHDA (help with plot_grid())](http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/)
