---
title: "Team Project Group 18"
author: "Nathan Patton, Reed Baumgardner, Fadumo Hussein"
output: 
 html_document:
  toc: true
  toc_depth: 3
  theme: lumen
  code_folding: show
---

# Executive Summary
One of the most unfortunate and disliked aspect of being an athlete is the risk of getting injured. Thousands of athletes are injured every single year, and these injuries vary with severity. In this report, the plan is to research how different variables effect the severity of injury in football. The end goal of this analysis is to lower the risk of injury for athletes, and make their lives healthier and easier, along with other stakeholders involved, such as family, teammates, athletic trainers, strength and conditioning coaches, etc. The established data set is detailed below, with emphasis on cleaning and sub-setting variables that are pertinent to the questions, along with a data dictionary that lists all the contents. To get a better understanding of the data sets contents, exploratory data analysis was performed, in addition to the data descriptions. In order to answer the questions of interest, detailed charts were created. Based on these charts, the key findings were.... 

# Report Dependencies
```{r message = FALSE}
library(tidyverse)
library(ggplot2)
library(mice)
library(skimr)
library(pollster)
library(kableExtra)
library(cowplot)
library(datadictionary)
library(pastecs)
```

# Establishing Data Set
[Kaggle-- NFL Injury Analysis](https://www.kaggle.com/code/aleksandradeis/nfl-injury-analysis/data)
```{r warning = FALSE}
#Loading in the data from local files
InjuryRecord <- read.csv("InjuryRecord.csv")
PlayList <- read.csv("PlayList.csv")

#Cleaning Data
##Checking for missing values injury record 
InjuryRecord[InjuryRecord==""] <- NA
colSums(is.na(InjuryRecord))
md.pattern(InjuryRecord, rotate.names = TRUE)

##Checking for missing values playlist 
PlayList[PlayList==""] <- NA
colSums(is.na(PlayList))
md.pattern(PlayList, rotate.names = TRUE)

##Joining PlayList with InjuryRecord 
injuries <- inner_join(PlayList, InjuryRecord, by="PlayKey")

##Removing duplicate columns
injuries <- select(injuries, -c("PlayerKey.y","GameID.y","Surface", "Position", "PositionGroup"))

##Renaming Columns
injuries <- injuries %>%
  rename("PlayerKey" = "PlayerKey.x", "GameID" = "GameID.x")

##Temperatures for indoor stadiums were incorrectly showing up as -999. We set it to 70 degrees based on the average on-field temperature NFL teams keep indoor stadiums. 
injuries$Temperature[injuries$Temperature == -999] <- 70

##Weather for indoor stadiums was set as "No Weather" for the purposes of this project
injuries$Weather[is.na(injuries$Weather) == TRUE] <- "No Weather"

##Random PlayerDay was changed from -34 to 34
injuries$PlayerDay[injuries$PlayerDay==-34] <- 34

##Combining some of the weather values
injuries$Weather[injuries$Weather==""|
                   injuries$Weather=="Indoors"|
                   injuries$Weather=="Controlled Climate"]<- "Indoor"
injuries$Weather[injuries$Weather=="Clear skies"|
                   injuries$Weather=="Clear and warm"|
                   injuries$Weather=="Clear Skies"]<- "Clear"
injuries$Weather[injuries$Weather=="Coudy"|
                   injuries$Weather=="Cloudy and Cool"|
                   injuries$Weather=="Cloudy, 50% change of rain"|
                   injuries$Weather=="Cloudy with periods of rain, thunder possible. Winds shifting to WNW, 10-20 mph."|
                   injuries$Weather=="Mostly cloudy"|
                   injuries$Weather=="Partly Cloudy"] <- "Cloudy"
injuries$Weather[injuries$Weather=="Light Rain"|
                   injuries$Weather=="Rain shower"]<- "Rain"
injuries$Weather[injuries$Weather=="Mostly Sunny"|
                   injuries$Weather=="Mostly sunny"|
                   injuries$Weather=="Sun & clouds"]<- "Sunny"

##Combining PlayType Variables
injuries$PlayType[injuries$PlayType=="Kickoff Not Returned"|
                    injuries$PlayType=="Kickoff Returned"] <- "Kickoff"
injuries$PlayType[injuries$PlayType=="Punt Not Returned"|
                    injuries$PlayType=="Punt Returned"]<- "Punt"
injuries <- na.omit(injuries)

##Forming a new days missed column using mutate 
injuries <- injuries %>%
  mutate(Days_Missed = ifelse(DM_M42 == 1, 42, ifelse(DM_M28 == 1, 28, ifelse(DM_M7 == 1, 7, 1))))

##Deleting the original days missed columns
injuries <- select(injuries, -c("DM_M1","DM_M7","DM_M28","DM_M42"))
View(injuries)

#Subsetting variables pertinent to questions and creating data dictionaries
question1data <- injuries %>%
  select(RosterPosition, BodyPart, StadiumType, Temperature, Weather, PlayType, Days_Missed, FieldType)
injury.labels <- c(RosterPosition = "This is the position the athlete has on the football team",
         BodyPart = "This variable records where the injury took place on the body",
         StadiumType = "The playing surface of the field when the injury occurred",
         Temperature = "Temperature recorded when the injury took place in Fahrenheit",
         Weather = "Type of weather record when the injury took place",
         PlayType = "Description of the play type when injury was recorded", 
         Days_Missed = "Measured the number of days missed due to injury", 
         FieldType = "Type of field where the injury took place")
     create_dictionary(question1data, var_labels =  injury.labels)
question2data <- question1data %>%
  filter(RosterPosition == "Wide Receiver")
injury.labels <- c(RosterPosition = "This is the position the athlete has on the football team",
         BodyPart = "This variable records where the injury took place on the body",
         StadiumType = "The playing surface of the field when the injury occurred",
         Temperature = "Temperature recorded when the injury took place in Fahrenheit",
         Weather = "Type of weather record when the injury took place",
         PlayType = "Description of the play type when injury was recorded", 
         Days_Missed = "Measured the number of days missed due to injury", 
         FieldType = "Type of field where the injury took place")
     create_dictionary(question2data, var_labels =  injury.labels)
```

#Background and Questions
## Background
This dataset was created by the National Football League (NFL) for a Kaggle-based analytics competition named "NFL 1st and Future - Analytics." The purpose of this competition was to "investigate the relationship between the playing surface and the injury and performance of NFL athletes." The inspiration for the competition was the recent investigations into the higher rate of injuries to NFL players on synthetic turf compared to that on natural turf. As a group, we were interested in sports and health, and this dataset was a reputable intersection of the two. In one of the nation's most profitable, yet dangerous, sports leagues, the risk of investment into a player can be lowered if steps continue to be taken to ensure player safety. While NFL player health is the foremost reason for the creation of this competition and dataset, we believe that conclusions from our analysis can be beneficial to the youth ranks of football as well. Keeping the children safe on the field was an important motivation for our analysis.  Our questions were stemmed from the NFL's own questions about safety. Our first question was an obvious one to come to. We wanted to explore all the variables in the dataset to find which ones were contributing most to injuries, especially severe ones that are career and sometimes life changing. Our second question was derived more specifically as we wanted to look into one of the NFL's most injury-prone positions-- Wide Receivers. This position group are some of the highest paid in the league and also one of the positions that are the most affected long-term by injuries as speed, agility, and sharpness are absolute keys to their success. These factors led us to focusing an analysis on just the wide receivers.

## Questions
1. Which conditions are linked to the most and most severe injuries? 
2. Which conditions are most/least linked to WR injuries?

# Descriptive Statistics and Exploratory Data Analysis
## Descriptive Statistics
### Seen below are a few descriptive statistic for the data necessary for answering question 1. There are five character variables, including RosterPosition, BodyPart, StadiumType, Weather and PlayType. Additionally, there are two numeric variables, inclduing temperature and Days_Missed. Temperature is a continuous predictor variable, which can be seen in the Exploratory Data Analysis section, and Days_Missed is the discrete outcome variable that is being analyzed. Some important takeaways from this output are the Min, Median, Mean, and Max of each of the numeric variables. These statistics will play a role in the future narratives.
```{r}
#Question 1 data (all positions)
summary(question1data)
```

### Seen below are a few descriptive statistic for the data necessary for answering question 2. The variables included are the same as for question one, but the only RosterPosition is wide receiver. Once again, some important takeaways from this output are the Min, Median, Mean, and Max of each of the numeric variables. Notice how values vary compared to the question 1 data. For example, the mean Days_Missed for the wide receiver postion is 10.93, whereas for all the positions combined it is almost twice as long at 17.96. This may suggest the wide receivers often miss less time when injured compared to some of the other roster positions.
```{r}
#Question 2 data (WR only)
summary(question2data)
```

## Exploratory Data Analysis
###Viewing Subsetted Variables Pertinent to Questions
```{r}
#1. Which conditions are linked to the most and most severe injuries? 
View(question1data)

#2. Which conditions are most/least linked to WR injuries?
View(question2data)
```

### Checking Distributions of the Continuous Variables
#### Question 1: The only continous variable within each of the datasets of interest is Temperature. A histogram of the variable can be seen below for the question 1 data. Take note of the wide range of temperatures as well as the most frequent temperature around 70 degrees fahrenheit. The data is slightly left skewed, but generally takes on a normal distribution. The mean is 65.34 which is relativelty close to the middle of the min, 33, and max 89 degrees fahrenheit.
```{r}
ggplot(question1data, aes(Temperature)) +
  geom_histogram(bins = 9, fill = '#540d6e') + 
  geom_vline(aes(xintercept = mean(Temperature)),col = 'lightgray', linewidth = 1) +
  geom_vline(aes(xintercept = median(Temperature)), col = 'lightgray', linetype = 'dashed') +
  labs(
    title = 'Histogram of Temperature',
    caption = 'mean: solid line, median = dashed line'
  ) +
  theme_minimal()
```

#### Question 2: A histogram of the Temperature variable can be seen below for the question 2 data. Once again, take note of the wide range of temperatures as well as the most frequent temperature around 70 degrees fahrenheit. The data is a normal distribution. The mean is 64.07 which is almost exactly in the middle of the min, 39, and max 89 degrees fahrenheit.
```{r}
ggplot(question2data, aes(Temperature)) +
  geom_histogram(bins = 4, fill = '#540d6e') + 
  geom_vline(aes(xintercept = mean(Temperature)),col = 'lightgray', linewidth = 1) +
  geom_vline(aes(xintercept = median(Temperature)), col = 'lightgray', linetype = 'dashed') +
  labs(
    title = 'Histogram of Temperature',
    caption = 'mean: solid line, median = dashed line'
  ) +
  theme_minimal()
```

# Question 1 Plots and Narrative
## Color Palette 
```{r}
pal = c("#8CB369","#2191FB","#F4A259","#BC4B51")
```

## Grouped Bar Chart for BodyPart by Days_Missed
```{r warning = FALSE}
ggplot(question1data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
    xlab("Body Part Injured") +
    ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for FieldType by Days_Missed
```{r warning = FALSE}
ggplot(question1data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Weather by Days_Missed
```{r warning = FALSE}
ggplot(question1data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather Conditions") +
  xlab("Weather Conditions") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Position by Days_Missed
```{r warning = FALSE}
ggplot(question1data, aes(x=RosterPosition, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Position") +
  xlab("Roster Position") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Play Type by Days_Missed
```{r warning = FALSE}
ggplot(question1data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Combined Chart
```{r warning = FALSE}
q1grid <- plot_grid(a1,a2,a3,a5,a4, 
          labels = c("A", "B", "C","D","E"),
          ncol = 2, nrow = 3)
q1grid
```
## Narrative 

#The count of days was higher on plays that took place on the synthetic field rather than natural fields. A play that took the place on cloudy days led to longer injury days. Plays that led to athletes having more missed days were passes. The position that has most days out due to injury, is linebacker. Finally, athletes who injured their knees were out the longest.  

#If our stakeholders aim to ensure the quality and health of athletes, we should be looking at potentially several solutions to resolve patterns observed in the bar graphs—the major one being rescheduling football games when conditions are cloudy. The next potential solution is playing on natural turf instead of synthetic turf.  Athletic trainers could reevaluate current conditioning plans to ensure athletes are not potentially over or under-training.  In regards to linebackers, we can ask to see if the NFL would create harsher penalties. For example, changing the unnecessary rule from a loss of 15 yards to 20 yards could potentially disincentivize athletes from doing rougher plays. 



# Question 2 Plots and Narrative
## Grouped Bar Chart for BodyPart by Days_Missed

```{r warning = FALSE}
b1<-ggplot(question2data, aes(x=BodyPart, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Body Part") +
  xlab("Body Part Injured") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for FieldType by Days_Missed
```{r warning = FALSE}
b2<-ggplot(question2data, aes(x=FieldType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Field Type") +
  xlab("Field Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Weather by Days_Missed
```{r warning = FALSE}
b3<-ggplot(question2data, aes(x=Weather, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Weather Conditions") +
  xlab("Weather Conditions") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Grouped Bar Chart for Play Type by Days_Missed
```{r warning = FALSE}
b4<-ggplot(question2data, aes(x=PlayType, y=..count.., fill=Days_Missed)) + 
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Days Missed by Play Type") +
  xlab("Play Type") +
  ylab("Count")+ 
  scale_fill_manual(values=pal)+
  theme_minimal()
```

## Combined Chart
```{r r warning = FALSE}
q2grid <- plot_grid(b1,b2,b3,b4, 
                    labels = c("A", "B", "C","D"),
                    ncol = 2, nrow = 2)
q2grid
```

## Narrative 

#We focused our next question on wide receivers. Once again we wanted to focus on this group of athletes due to How injury prone this role is. Injuries to the ankle cause these athletes to sit out for the longest.  there was no difference in field types.  however similar to the first set of combined graphs,  play that took place in cloudy weather led to longer injury days. Finally, plays that were a pass also led to wide receivers being out the longest. 

As stayed similarly before knowing our stakeholders and their goal to protect the health and wealth of our players here are potential solutions for wide receivers. The first would be to reschedule football games if the weather is cloudy.  another solution can be creating harsher penalties.  Finally, we can be looking at Having wide receivers wear ankle braces to ensure the stability of their ankles. However, we do recognize that adding any potential weight to them will slow their speed down.


# Summary

Our goal in this report is to ultimately give insight as to how we can ensure the safety of athletes.  This was our intent in each step of this report whether it be data cleaning looking at data descriptives and building plots that can give us insight but also potential solutions on how to keep athletes safe. A major finding was cloudy days and passes both had very high days missed. Injuries were high as well however varied based on position. Overall looking at the data we feel confident that we were able to answer both questions and additionally provide potential solutions to limit the risk of injury for players in the NFL.

# Citations
[STHDA (help with plot_grid())](http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/)

# Contrubutions:
Nathan Patton - Renaming columns and merging tables, cleaning, mutating days missed columns into one, and the first plot
Reed Baumgardner - Merging tables, removing duplicate columns and renaming columns, cleaning, additional comments on process
Fadumo Hussein - Find and relabel missing data, collaping variable factors, additional comments on process
