---
title: "Team Project Group 18"
output: 
 html_document:
  toc: true
  toc_depth: 3
  theme: lumen
  code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Group Names
Reed Baumgardner
Nathan Patton
Fadumo Hussein

# Topic
NFL Injuries

# Possible Data Sources
[Kaggle-- NFL Injury Analysis](https://www.kaggle.com/code/aleksandradeis/nfl-injury-analysis/data)

# Possible Questions
What variables correlate positively with NFL injuries? (injury surface, weather, stadium type, position, etc)

Is it more common to get injured in the cold/rain? Indoor/Outdoor stadiums?

Which variables are causing the most severe injuries (measured by days out)?

Which position is most at risk? In which conditions?

Analysis per position. Which variables are linked to the most QB injuries?

#Final Questions
1. Which conditions are linked to the most and most severe injuries? 
2. Which conditions are most/least linked to QB injuries?

#Packages 
```{r}
library(tidyverse)
library(ggplot2)
library(mice)
library(skimr)
library(pollster)
library(kableExtra)
```

#Uplaod Data 
```{r}
InjuryRecord <- read.csv("InjuryRecord.csv")
PlayList <- read.csv("PlayList.csv")
```

#Checking for missing values and cleaning up data 
```{r}
#Checking for missing values injury record 
InjuryRecord[InjuryRecord==""] <- NA
colSums(is.na(InjuryRecord))
md.pattern(InjuryRecord, rotate.names = TRUE)

#Checking for missing values playlist 
PlayList[PlayList==""] <- NA
colSums(is.na(PlayList))
md.pattern(PlayList, rotate.names = TRUE)
```

#Rename columns and merging table 
```{r}
#Joining PlayList with InjuryRecord 
injuries <- inner_join(PlayList, InjuryRecord, by="PlayKey")

#Removing duplicate columns
injuries <- select(injuries, -c("PlayerKey.y","GameID.y","Surface", "Position", "PositionGroup"))

#Renaming Columns
injuries <- injuries %>%
  rename("PlayerKey" = "PlayerKey.x", "GameID" = "GameID.x")
```

#More Cleaning
```{r}
# Temperatures for indoor stadiums were incrrectly showinf up as -999. We set it to 70 degrees based on the average on-field temperature NFL teams keep indoor stadiums. 
injuries$Temperature[injuries$Temperature == -999] <- 70
# Weather for indoor stadiums was set as "No Weather" for the purposes of this project
injuries$Weather[is.na(injuries$Weather) == TRUE] <- "No Weather"
View(injuries)

#Combine weather factors into n = 5 variables 

table(injuries$Weather)
injuries$Weather <- fct_collapse(injuries$Weather, 
                                 Clear =c("Clear", "Clear and warm", "Clear skies", "Clear Skies", "Sunny", "Mostly sunny", "Mostly Sunny"),
                                 Cloudy =c("Cloudy", "Cloudy and Cool", "Cloudy, 50% change of rain", "Coudy", "Partly Cloudy", "Sun & clouds", "Mostly cloudy"), 
                                 Rain =c("Cloudy with periods of rain, thunder possible. Winds shifting to WNW, 10-20 mph.", "Light Rain", "Rain", "Rain shower", "Rain", "Rain"), 
                                 
                                 Other =c("No Weather", "Indoor", "Indoors", "Controlled Climate", "Fair"))
table(injuries$Weather)
```

#Merging days missed columns into one outcome column
```{r}
#Forming a new days missed column using mutate 
injuries <- injuries %>%
  mutate(Days_Missed = ifelse(DM_M42 == 1, 42, ifelse(DM_M28 == 1, 28, ifelse(DM_M7 == 1, 7, 1))))

#Deleting the original days missed columns
injuries <- select(injuries, -c("DM_M1","DM_M7","DM_M28","DM_M42"))
View(injuries)
```

#First Plot
```{r}
ggplot(injuries,               
       aes(x = Temperature,           
           y = Days_Missed,        
           color = RosterPosition)) +   
  geom_point(aes(shape = FieldType))
```
#Contrubutions:
Nathan Patton - Renaming columns and merging tables, cleaning, mutating days missed columns into one, and the first plot
Reed Baumgardner - Merging tables, removing duplicate columns and renaming columns, cleaning, additional comments on process
Fadumo Hussein - Find and relabel missing data, collaping variable factors, additional comments on process




#Exploratory Data Analysis
##Subsetting Variables Pertinent to Questions
###1. Which conditions are linked to the most and most severe injuries?
```{r}
question1data <- injuries %>%
  select(RosterPosition, StadiumType, FieldType, Temperature, Weather, PlayType, DM_M1, DM_M7, DM_M28, DM_M42)
skim(injuries)
View(question1data)
```
##Checking Distributions of Continuous Variables
###Outcome Variables - Number of Games Missed (Need to do some editing and pivot longer before doing this)
###Predictor Variables
####Temperature
```{r}
ggplot(injuries, aes(Temperature)) +
  geom_histogram(bins = 200, fill = '#540d6e') + 
  geom_vline(aes(xintercept = mean(Temperature)),col = 'lightgray', linewidth = 1) +
  geom_vline(aes(xintercept = median(Temperature)), col = 'lightgray', linetype = 'dashed') +
  labs(
    title = 'Histogram of Temperature',
    caption = 'mean: solid line, median = dashed line'
  ) +
  theme_minimal()
```

##Checking for Outliers in Coninuous Variables
```{r}
ggplot(injuries, aes(Temperature)) +
  geom_boxplot(fill = '#540d6e') + 
  labs(
    title = 'Boxplot of Temperature'
  ) +
  theme_minimal()
```
THESE GRAPHS ARE NOT GREAT BECAUSE WE STILL NEED TO DO SOME DATA CLEANING
- WE WERE WORKING ON THAT MOST OF CLASS
